---
title: "Healthcare Needs Report"
author: "Sudeep Santhosh"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


```{r}
library(lme4)
library(tidyverse)
library(here)
library(haven)
library(labelled)
library(sjlabelled)
library(ggplot2)
library(dplyr)
library(sjPlot)
library(naniar)
library(tinytable)
library(ggmosaic)
options(tinytable_tt_digits = 3)
options(tinytable_latex_placement = "H")
```

# Data loading
The `.sav` format was selected for data loading because it preserves variable labels, data types, and metadata more effectively than flat file formats such as `.csv`. This is particularly useful for analytically work.
```{r}
healthy_wave=haven::read_sav(here("data","CRONOS3 Theme - Make it Healthy.sav"))
dictionary=labelled::generate_dictionary(healthy_wave)
saveRDS(healthy_wave, here("data", "healthy_wave.rds"))
```

# Exploratory Data Analysis (EDA)

## Duplicate value 

```{r}
sum(duplicated(healthy_wave$idno))
```

```{r}

healthy_wave$needed_any=ifelse(
  rowSums(healthy_wave[, c("w1hq1", "w1hq3", "w1hq6", "w1hq11")] == 1, na.rm = TRUE) > 0,
  1, 0
)
```

```{r}
healthy_wave %>%
  filter(idno==25213)|>
  select(cntry,age,idno)|>
  haven::as_factor() |>
  tt()
```
Checking id-25213  

In `healthy_wave1`, 25213 id has two entries with countries as CZ and FI.

Trying to see how the ids are filled, realized that for every country there is a given range of values assigned.

So as to create unique values, we will be merging `cntry` and `idno` to create a new value, which will be the new primary key for filtering.

Solution-1
solving duplicates by merging the `cntry` and `idno` column 

```{r}
primary_health_wave=healthy_wave|>
  mutate(new_id=paste(cntry,idno,sep= "_"))
sum(duplicated(primary_health_wave$new_id))
```
still 21 duplicate values exist so we merge another column which is `age` to create a primary key in the dataset
also to check if its not just the same person and not at different ages / years we had to check this 


```{r}
primary_health_wave|>
  filter(duplicated(new_id) | duplicated(new_id,fromLast=TRUE)) |>
  arrange(new_id)|>
  select(cntry, new_id,age,gndr)|>
  haven::as_factor()|>
  slice_head(n=12)|>
  tt()
```
from this we can say that none of them are the same person as there huge again difference gap of more  than one year 

on merging with with three columns which is  `age` ,`idno` and `cntry` we were able to remove the duplicates so now we can use this column as a primary key for identifying an individual

```{r}
primary_health_wave=primary_health_wave|>
  mutate(new_id=paste(cntry,idno,age, sep="_"))
sum(duplicated(primary_health_wave$new_id))
```



## Model parametrs at intial phase

 no of people who Needed to see a general practitioner last 12 months but could not 
```{r}
sum(healthy_wave$w1hq1 ==1,na.rm = TRUE)
```
no fo people who Needed to see dentist last 12 months but could not
```{r}
sum(healthy_wave$w1hq6 ==1,na.rm = TRUE)
```
no of people Needed to see medical specialists but could not 
```{r}
sum(healthy_wave$w1hq3 == 1, na.rm = TRUE)
```

```{r}
length(healthy_wave$idno[
  which(healthy_wave$w1hq6==1|
        healthy_wave$w1hq3==1|
        healthy_wave$w1hq1==1|
        healthy_wave$w1hq11==1)
])
```

we have increased the number of observation to 1048

### Variable Description Table

| Variable | Role | Description |
|----------|------|-------------|
| w1hq1 | Response | Needed to see a general practitioner in the last 12 months |
| w1hq3 | Response | Needed to see medical specialists in the last 12 months |
| w1hq6 | Response | Needed to see dentist in the last 12 months |
| w1hq11 | Response | Needed diagnostic tests or scans prescribed by a doctor in the last 12 months |
| w1hq5 | Predictor | Wait time for medical specialist appointment |
| cntry | Predictor | Country |
| age | Predictor | Age |
| gndr | Predictor | Gender |
| w3hq77 | Predictor | Currently in paid work |
| w3hq57 | Predictor | Subjective mental health in general |
| w3hq58 | Predictor | Subjective physical health in general |
| w3hq87 | Predictor | Difficulty taking an hour off for personal/family matters during work |
| w5hq18 | Predictor | Healthcare costs covered by system |
| w5hq11 | Predictor | Receive help and support from close contacts when needed |
| w5hq8 | Predictor | Severe financial difficulties in family during first 18 years of life |
| w5hq17 | Predictor | Out-of-pocket cost for most recent private mental health consultation |
| w5hq12 | Predictor | Felt lonely during the past week |



## Country wise analysis
```{r}
country_summary=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    total_people=n(),
    people_needed=sum(needed_any, na.rm = TRUE),
    proportion=people_needed/total_people
  )|>
  haven::as_factor() |>
  tt()
country_summary
```

```{r, echo=FALSE}
primary_health_wave$needed_any=factor(
  primary_health_wave$needed_any,
  levels = c(0, 1),
  labels = c("Met Health Care", "Unmet Health Care")
)
```

```{r, echo=FALSE}
primary_health_wave$gndr=factor(
  primary_health_wave$gndr,
  levels = c(1, 2),
  labels = c("Male", "Female")
)
```

```{r}
ggplot(primary_health_wave)+
  geom_mosaic(aes(x=product(needed_any,cntry),
                  fill=needed_any))+
  scale_x_productlist("Country") +
  scale_y_productlist("Medical Care Status") +
  scale_fill_manual(
    values = c("orange", "blue"),
    name = "Medical Care Status"
  ) +
  theme_minimal()+
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```


The mosaic plot suggests that, across all countries, most respondents reported having their medical care needs met. However, the proportion of unmet health care need varies by country, with relatively higher shares observed in countries such as **GB(United Kingdom)**, **PL(Poland)**, **PT(Portugal)**, and **FI(Finland)**. This indicates potential cross-country differences in access to or delivery of medical care services.
```{r}
country_summary1=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    people_needed=sum(needed_any == "Unmet Health Care", na.rm = TRUE),
  )|>
  arrange(desc(people_needed))|>
  haven::as_factor() |>
  tt()
  

country_summary1
```

```{r}
service_country=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm=TRUE),
    Dentist=sum(w1hq6==1,na.rm=TRUE),
    Diagnostic_Test=sum(w1hq11==1,na.rm=TRUE))|>
  pivot_longer(
    cols=-cntry,
    names_to="Service",
    values_to="Count")
ggplot(service_country,aes(x=reorder(cntry,-Count),y=Count,fill=Service))+
  geom_bar(stat="identity")+
  labs(title="Unmet Healthcare Services by Country",
    x="Country",
    y="Count")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45,hjust=1))
```




## Variable Analysis thorugh likert charts 
| Variable | Description |
|----------|-------------|
| w3hq57 | Subjective mental health in general |
| w3hq58 | Subjective physical health in general |
| w5hq12 | Felt lonely, how much of the time in the past week |

```{r}
label_summary <- tibble(
  Variable = c(
    "Subjective mental health in general",
    "Subjective physical health in general",
    "Felt lonely, how much of the time past week"
  ),
  Labels = c(
    paste(names(attr(healthy_wave$w3hq57, "labels")), collapse = ", "),
    paste(names(attr(healthy_wave$w3hq58, "labels")), collapse = ", "),
    paste(names(attr(healthy_wave$w5hq12, "labels")), collapse = ", ")
  )
)

tt(label_summary)
```


The labels Very Positive, Positive, Decent, Negative, and Very Negative were chosen to create a consistent ordinal framework across the three well being variables: `loneliness`, `subjective mental health`, and `subjective physical health`. Although the original survey items use slightly different response wording (for example, frequency-based responses for loneliness and quality-based responses for health status), recoding them into a common directional scale improves the comparability and interpret ability of the Likert chart.

```{r}
primary_health_wave$lonely=factor(primary_health_wave$w5hq12,levels = c(1, 2, 3, 4),labels = c("Very Positive","Positive","Decent","Negative" ),ordered = TRUE)

primary_health_wave$mental_state=factor(primary_health_wave$w3hq57,levels = c(1, 2, 3, 4,5),labels = c("Very Positive","Positive","Decent","Negative","Very Negative"),ordered = TRUE)
primary_health_wave$physical_state=factor(primary_health_wave$w3hq58,levels = c(1, 2, 3, 4,5),labels = c("Very Positive","Positive","Decent","Negative","Very Negative"),ordered = TRUE)

df_yes=primary_health_wave[primary_health_wave$needed_any== "Unmet Health Care", ]
df_no=primary_health_wave[primary_health_wave$needed_any=="Met Health Care", ]
```


```{r}
plot_likert(df_yes[,c("mental_state", "physical_state","lonely")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
show.prc.sign = TRUE,geom.colors = "RdBu")
```

```{r}
plot_likert(df_no[,c("mental_state", "physical_state","lonely")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
show.prc.sign = TRUE,geom.colors = "RdBu")
```
From this we can say that mental state of people who did receive medical care they had bad mental state as well as a bad physical state which makes it a good parameter for the model, we can also say loneliness is kind of high for people who need health care 


## Missing Data Analysis

| Variable Code | Description |
|--------------|-------------|
| w3hq87 | How difficult to take an hour off for personal/family matters during work |
| w5hq18 | Healthcare costs covered by system |
| w5hq11 | Receive help and support from close contacts when needed |
| w5hq8 | Severe financial difficulties in family during first 18 years of life, how often |
| w5hq17 | Out-of-pocket cost for most recent private mental health consultation |
| w5hq12 | Felt lonely, how much of the time past week |
| w3hq77 | Currently in paid work |


```{r}
missing_raw=miss_var_summary(
  df_yes[, c("w3hq87", "w5hq18", "w5hq11", "w5hq8", "w5hq17", "w5hq12","w3hq77")]
)
missing_table=tibble(
  Variable=c(
    "Time off work difficulty",
    "Healthcare cost coverage",
    "Social support",
    "Childhood financial difficulty",
    "Mental health consultation cost",
    "Loneliness",
    "Currently in paid work"
  ),
  Missing_Count=as.numeric(missing_raw$n_miss),
  Missing_Percentage=as.numeric(missing_raw$pct_miss)
)
tt(missing_table)
```




we might need to remove two varible for the model to take btter resutls 
w5hq17- Out-of-pocket cost for most recent private mental health consultation -86.6%
w3hq87- How difficult to take an hour off for personal/family matters during work -50.7% 
as they very high missing values so its better to remove them 


```{r}
gg_miss_upset(primary_health_wave[,c("w3hq87", "w5hq18","w5hq11","w5hq8","w5hq17","w5hq12","w3hq77")])
```



removing both those column remove a high chunk of missing values as we can see `w5hq17` and a combination of both `w3hq87` with `w5hq17` have the most missing values 



```{r}
gg_miss_case(primary_health_wave[,c("w3hq87", "w5hq18","w5hq11","w5hq8","w5hq17","w5hq12","w3hq77")])
```
comparing the miss cases on removing the two columns 
```{r}
gg_miss_case(primary_health_wave[,c( "w5hq18","w5hq11","w5hq8","w5hq12","w3hq77")])
```



```{r}
ggplot(primary_health_wave,aes(x=factor(needed_any),y=age,fill=factor(gndr)))+
  geom_boxplot(outlier.shape = NA)+ 
  labs(x="Healthcare Need Status",
    y="Age",
    title="Age Distribution by Healthcare Need Status")+
  scale_x_discrete(labels=c("Met","Unmet"))+
  scale_fill_manual(name="Gender",
    labels=c("Male", "Female","No answer"),
    values=c("blue","pink","black")
  )+facet_wrap(vars(cntry))
```



```{r}
primary_health_wave |>
  filter(!is.na(gndr))|>
  group_by(cntry, needed_any, gndr)|>
  summarise(
    n = n(),
    median_age = median(age, na.rm = TRUE),
    age_range = paste(
      quantile(age, 0.25, na.rm = TRUE),
      " - ",
      quantile(age, 0.75, na.rm = TRUE)
    ),.groups = "drop")|>
  haven::as_factor() |>
  tt()
```




Across most countries, the median ages for respondents with met and unmet healthcare needs are broadly similar, though some notable differences emerge.

- **Ireland:** unmet males = 59.5, met males = 57  
- **Austria:** unmet females = 56.5, met females = 53  
- **Poland:** unmet females = 49.5, met females = 45  

This suggests that in some countries, those reporting unmet healthcare needs tend to be slightly older. However, this is not universal.

- **Iceland:** unmet females = 42 vs met females = 54  
- **Slovenia:** unmet males = 40 vs met males = 52  

Across nearly all countries, female respondents generally have slightly lower median ages than males within the same healthcare need group. This pattern is visible in both the boxplot medians and the table.

The interquartile ranges (`age_range`) show that age dispersion differs across countries.

- **Austria** females unmet: 36–73.25 — wide spread  
- **Poland** females unmet: 36.5–68.25 — wide spread  
- **Iceland** females unmet: 31–52 — much narrower


$$
\begin{aligned}
\log \left(
\frac{P(\text{Unmet Health Care Need})}
{1 - P(\text{Unmet Health Care Need})}
\right)
&= \beta_0 + \beta_1(\text{Age}) + \beta_2(\text{Gender}) + \beta_3(\text{Country}) \\
&\quad + \beta_4(\text{Subjective Mental Health}) + \beta_5(\text{Subjective Physical Health}) \\
&\quad + \beta_6(\text{Loneliness}) + \beta_7(\text{Social Support}) \\
&\quad + \beta_8(\text{Currently in Paid Work}) \\
&\quad + \beta_9(\text{Early-Life Financial Difficulty}) \\
&\quad + \beta_{10}(\text{Healthcare Costs Covered})
\end{aligned}
$$



```{r}
model_0=glm(
  needed_any ~ age + gndr + cntry +
    w3hq57 + w3hq58 +w5hq12 + w5hq11 +w3hq77+w5hq8 + w5hq18 ,
  family = binomial,
  data = primary_health_wave
)

summary(model_0)
```



Based on the output we can say that 

formula used 


$$
OR = e^{\beta}
$$

- As the estimate of age is negative each additional year is associated with **0.6% lower odds** of unmet healthcare need and is significantly associated with unmet healthcare need

- Although females show slightly higher estimated odds, this difference is not statistically significant

- Substantial cross-country differences were observed in the odds of reporting unmet healthcare need relative to Austria. Belgium exhibited significantly lower odds, while most other countries showed significantly higher odds. Particularly large effects were observed for the United Kingdom, Poland, and Portugal, where respondents had more than twice the odds of reporting unmet healthcare need compared with Austria. Hungary showed higher estimated odds; however, this difference was not statistically significant.

- Worse subjective mental health was associated with slightly higher estimated odds of reporting unmet healthcare need (approximately 5% per one-level increase); however, this relationship was not statistically significant.

- Worse subjective physical health was significantly associated with unmet healthcare need. Specifically, each one-level worsening in physical health was associated with approximately **19% higher odds** of reporting unmet healthcare need.next 

- Higher levels of loneliness were significantly associated with unmet healthcare need. Specifically, each one-level increase in loneliness was associated with approximately **29% higher odds** of reporting unmet healthcare need.

- Greater social support from close contacts was associated with a very small decrease in the estimated odds of unmet healthcare need; however, this relationship was not statistically significant.

- Being in paid work was associated with approximately **9% higher** estimated odds of reporting unmet healthcare need; however, this relationship was not statistically significant.

- Lower levels of childhood financial difficulty were significantly associated with reduced odds of reporting unmet healthcare need. Specifically, each one-level movement toward less frequent financial hardship during the first 18 years of life was associated with approximately **15% lower odds** of unmet healthcare need.

- Poorer perceived healthcare cost coverage was strongly and significantly associated with unmet healthcare need. Specifically, each one-level movement toward disagreement that healthcare costs are covered by the system was associated with approximately **34% higher odds** of reporting unmet healthcare need.


Drawbacks

- ordinal variables were not at level we are alsoing having gender baseline cntry basline and so on 

making separate data sets for the orindal varaible so we can anyalyse based on anova for later understadning 

```{r}
PHW=primary_health_wave
```


```{r}
PHW$w3hq77=factor(
  primary_health_wave$w3hq77,
  levels = c(1,2),
  labels = c("Yes","No")
)
```


```{r}
PHW$w3hq57=factor(
  primary_health_wave$w3hq57,
  levels=c(5,4,3,2,1),
  labels=c("Very bad","Bad","Fair","Good","Very good"),
  ordered=TRUE
)
```


```{r}
PHW$w3hq58=factor(
  primary_health_wave$w3hq58,
  levels=c(5,4,3,2,1),
  labels=c("Very bad","Bad","Fair","Good","Very good"),
  ordered=TRUE
)
```

```{r}
PHW$w5hq8=factor(
  primary_health_wave$w5hq8,
  levels=c(5,4,3,2,1),
  labels=c("Never","Hardly ever","Sometimes","Often","Always"),
  ordered=TRUE
)
```

```{r}
PHW$w5hq12=factor(
  primary_health_wave$w5hq12,
  levels = c(1,2,3,4),
  labels = c(
    "None or almost none of the time",
    "Some of the time",
    "Most of the time",
    "All or almost all of the time"
  ),
  ordered = TRUE
)
```

```{r}
PHW$w5hq18=factor(
  primary_health_wave$w5hq18,
  levels = c(5,4,3,2,1),
  labels = c(
    "Disagree strongly",
    "Disagree",
    "Neither agree nor disagree",
    "Agree",
    "Agree strongly"
  ),
  ordered = TRUE
)
```

Model 1 having level for your orindal varibles
```{r}
model_1=glm(
  needed_any ~ age + gndr + cntry +
    w3hq57 + w3hq58 +w5hq12 + w5hq11 +w3hq77+w5hq8 + w5hq18 ,
  family = binomial,
  data = PHW
)

summary(model_1)
```

- There is weak evidence suggesting that age may be negatively associated with unment health care but the effect is not statistically significant at the 5% level

- Females have about 14% higher odds of having unmet healthcare needs compared to males, holding all other variables constant but the effect is statistically significant at 5% singificanat level 

- Country of residence shows substantial variation in unmet healthcare needs. Compared to the reference  `AT` (Austria), individuals in several countries—such as `GB`(United Kingdom), `PL`(Poland), `PT`(Portugal), `FI`(Finland), `FR`(France), and `SI`(Slovenia) —have significantly higher odds of reporting unmet healthcare needs. For example, individuals in `GB` have nearly 2.8 times higher odds compared to those in `AT`.In contrast, individuals in `BE` have significantly lower odds of unmet healthcare needs relative to `AT`. Some countries (e.g., `HU`(Hungary)) do not show statistically significant differences from the reference category, indicating that their levels of unmet healthcare needs are not distinguishable from `AT` based on this model.


- `w3hq58` There is a significant negative linear relationship between Subjective physical health in general and unmet health care

-  `w5hq12`There is a significant positive linear relationship between Felt lonely during the past week parameter and unmet health care

- `w5hq8` There is a significant positive linear relationship between financial difficulties bellow the age of 18 parameter and unmet health care 

- `w5hq18` Healthcare costs covered by system and unmet healthcare shows a strong and statistically significant negative linear association  In addition, significant higher-order (cubic and quadratic) terms suggest that the relationship is not strictly linear, indicating some complexity in how different levels of the variable relate to the outcome





Mixed-Effects Logistic Regression Model

$$
\begin{aligned}
\log \left(
\frac{P(\text{Unmet Health Care Need}_{ij})}
{1 - P(\text{Unmet Health Care Need}_{ij})}
\right)
&= \beta_0 + u_j + \beta_1(\text{Age}_{ij}) + \beta_2(\text{Gender}_{ij}) \\
&\quad + \beta_3(\text{Subjective Mental Health}_{ij}) \\
&\quad + \beta_4(\text{Subjective Physical Health}_{ij}) \\
&\quad + \beta_5(\text{Loneliness}_{ij}) \\
&\quad + \beta_6(\text{Social Support}_{ij}) \\
&\quad + \beta_7(\text{Currently in Paid Work}_{ij}) \\
&\quad + \beta_8(\text{Early-Life Financial Difficulty}_{ij}) \\
&\quad + \beta_9(\text{Healthcare Costs Covered}_{ij})
\end{aligned}
$$

where:

$$
u_j \sim N(0, \sigma^2_{\text{country}})
$$
```{r}
model_5=glmer(
  needed_any~age+gndr+w5hq18+w5hq12+w5hq8+w3hq58+w3hq57+w5hq11+w3hq77+(1|cntry),
  family=binomial,
  data=primary_health_wave
)
summary(model_5)

```

- `w3hq58` There is a significant negative linear relationship between Subjective physical health in general and unmet health care after accounting for country-level variation

- `w5hq12` There is a significant positive linear relationship between Felt lonely during the past week parameter and unmet health care after accounting for country-level variation

- `w5hq8` There is a significant positive linear relationship between financial difficulties below the age of 18 parameter and unmet health care after accounting for country-level variation

- `w5hq18` Healthcare costs covered by system and unmet healthcare shows a strong and statistically significant 

- `w3hq57` There is no statistically significant relationship between Subjective emotional health in general and unmet health care after accounting for country-level variation

- `w5hq11` There is no statistically significant relationship between this parameter and unmet health care after accounting for country-level variation

- `w3hq77` There is no statistically significant relationship between this parameter and unmet health care after accounting for country-level variation

- `age` There is a weak negative relationship between age and unmet health care, however this is only marginally significant after accounting for country-level variation

- `gndr` There is no statistically significant relationship between gender and unmet health care after accounting for country-level variation


making model now without levels 
```{r}
model_6=glmer(
  needed_any~age+gndr+w5hq18+w5hq12+w5hq8+w3hq58+w3hq57+w5hq11+w3hq77+(1|cntry),
  family=binomial,
  data=PHW
)
summary(model_6)
```


```{r}

comparison_table <- data.frame(
  Model = c("Mixed-effects logistic regression with numeric encoding",
            "Mixed-effects logistic regression with ordered factor encoding",
            "Logistic regression with numeric encoding",
            "Logistic regression with ordered factor encodin"),
  AIC = c(AIC(model_5), AIC(model_6),AIC(model_0),AIC(model_1)),
  BIC = c(BIC(model_5), BIC(model_6),BIC(model_0),BIC(model_1)),
  LogLik = c(
    as.numeric(logLik(model_5)),
    as.numeric(logLik(model_6)),
    as.numeric(logLik(model_0)),
    as.numeric(logLik(model_1))
    
  )
)
tt(comparison_table)
```
Although ordered factor encoding improved model fit as indicated by higher log-likelihood values, the substantial increase in model complexity resulted in higher BIC values. Numeric encoding provided a more parsimonious model and achieved the lowest AIC among all candidate models.


Interactions between variables 

possible interaction between variables and why 

1) age * gndr
An interaction between age and gender was explored to investigate whether the relationship between age and unmet healthcare need differed between males and females. This interaction was theoretically justified, as healthcare utilisation patterns and barriers to care often vary across both age and gender groups. Exploratory visualisation of the data further suggested potential differences in age distributions between males and females across met and unmet healthcare need categories in several countries, supporting the inclusion of the interaction term in the mixed-effects logistic regression model

2)  loneliness * physical_health
The interaction between loneliness and subjective physical health was explored to assess whether the relationship between physical health and unmet healthcare need differed according to levels of social isolation. Individuals experiencing poor physical health may face greater barriers to healthcare access when combined with loneliness, as limited social support and increased isolation can further restrict healthcare-seeking behaviour. Therefore, the interaction term was included to examine the combined effect of physical health and loneliness on unmet healthcare need.


posssible interaction but will need diagrams to prove this for reasoning 



