---
title: "Untitled"
author: "Sudeep Santhosh"
date: "2026-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
library(haven)
library(labelled)
library(sjlabelled)
library(ggplot2)
library(dplyr)
library(sjPlot)
library(naniar)
```

loading all the waves data
```{r}
healthy_wave=haven::read_sav(here("data","CRONOS3 Theme - Make it Healthy.sav"))
dictionary=labelled::generate_dictionary(healthy_wave)
```

checking for duplicate values 
```{r}
sum(duplicated(healthy_wave$idno))
```

 no of people who Needed to see a general practitioner last 12 months but could not 
```{r}
sum(healthy_wave$w1hq1 ==1,na.rm = TRUE)
```
no fo people who Needed to see dentist last 12 months but could not
```{r}
sum(healthy_wave$w1hq6 ==1,na.rm = TRUE)
```
no of people Needed to see medical specialists but could not 
```{r}
sum(healthy_wave$w1hq3 == 1, na.rm = TRUE)
```

```{r}
length(healthy_wave$idno[
  which(healthy_wave$w1hq6==1|
        healthy_wave$w1hq3==1|
        healthy_wave$w1hq1==1|
        healthy_wave$w1hq11==1)
])
```

we have increased the number of observation to 1048

unmet healthcare need
merged Predictor variable required 
dataset           Response variable
healthy_wave     (w1hq1)   Needed to see a general practitioner last 12 months
healthy_wave     (w1hq3)   Needed to see medical specialists last 12 months 
healthy_wave     (w1hq6)   Needed to see dentist last 12 months
healthy_wave      (w1hq11)  Needed diagnostic tests or scans, prescribed by a doctor last 12

dataset            predicotr variable
healthy_wave      (w1hq5)   Wait time for medical specialist appointment 
""                (cntry)   Country
""                (age)     age

healthy_wave     (w3hq77)  Currently in paid work
""                (w3hq57)  Subjective mental health in general
""                (w3hq58)  Subjective physical health in general
""                (w3hq87)  How difficult to take an hour off for personal/family matters during work

healthy_wave     (w5hq18)  Healthcare costs covered by system
""                (w5hq11)  Receive help and support from close contacts when needed
""                (w5hq8)   Severe financial difficulties in family during first 18 years of life, how often
""                (w5hq17)  Out-of-pocket cost for most recent private mental health consultation
""                (w5hq12)  Felt lonely, how much of the time past week


handling duplicate values 

```{r}

healthy_wave$needed_any=ifelse(
  rowSums(primary_health_wave[, c("w1hq1", "w1hq3", "w1hq6", "w1hq11")] == 1, na.rm = TRUE) > 0,
  1, 0
)

```
there are are a total 
going to see how to handle duplicate values based on 
checking id-25213 
in healthy_wave1 25213 id has two entries with countries as CZ and FI 
trying to see how the ids are filled realized that for every country that a given range of values to assigned 
so as to create unique values we will be merging cntry and idno to create new value which will be the new primary key for filtering 

solving duplicates by merging the cntry and idno column 
checking with healthwave_1 dataset 

```{r}
primary_health_wave=healthy_wave|>
  mutate(new_id = paste(cntry, idno, sep = "_"))
sum(duplicated(primary_health_wave$new_id))
```
on merging a column country and idno to create a new unque id there still exist duplicates but has got reduced 1431 to 21 

```{r}
primary_health_wave=primary_health_wave|>
  mutate(new_id = paste(cntry,idno,age, sep = "_"))
sum(duplicated(primary_health_wave$new_id))
```
removed all duplicates

Country wise analysis
```{r}
country_summary=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    total_people = n(),
    people_needed = sum(needed_any, na.rm = TRUE),
    proportion = people_needed/total_people
  )
country_summary
```

```{r}
ggplot(country_summary, aes(x = reorder(cntry, -people_needed), 
                            y = people_needed)) +
  geom_bar(stat = "identity") +
  labs(
    title = "amount of People Needing Healthcare by Country",
    x = "Country",
    y = "amaount"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}

country_summary1=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    people_needed = sum(needed_any, na.rm = TRUE),
  )|>
  arrange(desc(people_needed))

country_summary1
```

```{r}
primary_health_wave |>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm = TRUE),
    dentist=sum(w1hq6==1,na.rm = TRUE),
    diagnostic_test_req=sum(w1hq11==1,na.rm = TRUE)
  )|>pivot_longer(
    cols=everything(),
    names_to = "Service",
    values_to = "Count"
  )|>
  ggplot(aes(x=Service,y=Count))+
  geom_bar(stat = "identity")+
  labs(
    title = " Unmet health care services across 11 countires",
    x = "Service Type",
    y = "Count"
  )+ scale_x_discrete(labels = c(
  GP = "General Practitioner",
  Specialist = "Specialist",
  Dentist = "Dentist",
  diagnostic_test_req = "Diagnostic Test"
))
```


```{r}
primary_health_wave|>
  filter(cntry=="GB")|>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm = TRUE),
    dentist=sum(w1hq6==1,na.rm = TRUE),
    diagnostic_test_req=sum(w1hq11==1,na.rm = TRUE)
  )|>pivot_longer(
    cols=everything(),
    names_to = "Service",
    values_to = "Count"
  )|>
  ggplot(aes(x=Service,y=Count))+
  geom_bar(stat = "identity")+
  labs(
    title = " Unmet health care services United Kingdom",
    x = "Service Type",
    y = "Count"
  )+ scale_x_discrete(labels = c(
  GP = "General Practitioner",
  Specialist = "Specialist",
  Dentist = "Dentist",
  diagnostic_test_req = "Diagnostic Test"
))
```

```{r}
primary_health_wave|>
  filter(cntry=="FI")|>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm = TRUE),
    dentist=sum(w1hq6==1,na.rm = TRUE),
    diagnostic_test_req=sum(w1hq11==1,na.rm = TRUE)
  )|>pivot_longer(
    cols=everything(),
    names_to = "Service",
    values_to = "Count"
  )|>
  ggplot(aes(x=Service,y=Count))+
  geom_bar(stat = "identity")+
  labs(
    title = " Unmet health care services Finland",
    x = "Service Type",
    y = "Count"
  )+ scale_x_discrete(labels = c(
  GP = "General Practitioner",
  Specialist = "Specialist",
  Dentist = "Dentist",
  diagnostic_test_req = "Diagnostic Test"
))
```

```{r}
primary_health_wave|>
  filter(cntry=="BE")|>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm = TRUE),
    dentist=sum(w1hq6==1,na.rm = TRUE),
    diagnostic_test_req=sum(w1hq11==1,na.rm = TRUE)
  )|>pivot_longer(
    cols=everything(),
    names_to = "Service",
    values_to = "Count"
  )|>
  ggplot(aes(x=Service,y=Count))+
  geom_bar(stat = "identity")+
  labs(
    title = " Unmet health care services Belgium",
    x = "Service Type",
    y = "Count"
  )+ scale_x_discrete(labels = c(
  GP = "General Practitioner",
  Specialist = "Specialist",
  Dentist = "Dentist",
  diagnostic_test_req = "Diagnostic Test"
))
```

likert scale analysis for the columns 
dataset            predicotr variable
healthy_wave3     
""                (w3hq57)  Subjective mental health in general
""                (w3hq58)  Subjective physical health in general
""                (w5hq12)  Felt lonely, how much of the time past week

first making a likert scale on merged_datset having only poeple with unmet halth care 




```{r}
primary_health_wave$mental_state=factor(primary_health_wave$w3hq57,levels = c(1, 2, 3, 4,5),labels = c("Very good","Good","Fair","Bad","Very bad"),ordered = TRUE)
primary_health_wave$physical_state=factor(primary_health_wave$w3hq58,levels = c(1, 2, 3, 4,5),labels = c("Very good","Good","Fair","Bad","Very bad" ),ordered = TRUE)


df_yes=primary_health_wave[primary_health_wave$needed_any== 1, ]
df_no=primary_health_wave[primary_health_wave$needed_any==0, ]
```


```{r}
plot_likert(df_yes[,c("mental_state", "physical_state")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
show.prc.sign = TRUE,geom.colors = "RdBu")
```

```{r}
plot_likert(df_no[,c("mental_state", "physical_state")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
geom.colors = "RdBu",
 show.prc.sign = TRUE)
```
missing data analysis on 
(w3hq87)  How difficult to take an hour off for personal/family matters during work
(w5hq18)  Healthcare costs covered by system
(w5hq11)  Receive help and support from close contacts when needed
(w5hq8)   Severe financial difficulties in family during first 18 years of life, how often
```{r}
miss_var_summary(df_yes[, c("w3hq87", "w5hq18","w5hq11","w5hq8")])
```
the largest percentage of missing data is in column  w3hq87 of 55%  and the rest shows 20% missing data this is for only people who have not revived health care 


```{r}
gg_miss_upset(df_yes[, c("w3hq87", "w5hq18","w5hq11","w5hq8")])
```
```{r}
gg_miss_upset(df_no[, c("w3hq87", "w5hq18","w5hq11","w5hq8")])
```
```{r}
gg_miss_case(df_yes)
```





















