---
title: "Untitled"
author: "Sudeep Santhosh"
date: "2026-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


```{r}
library(tidyverse)
library(here)
library(haven)
library(labelled)
library(sjlabelled)
library(ggplot2)
library(dplyr)
library(sjPlot)
library(naniar)
library(tinytable)
library(ggmosaic)
options(tinytable_tt_digits = 3)
options(tinytable_latex_placement = "H")
```

# Data loading
The `.sav` format was selected for data loading because it preserves variable labels, data types, and metadata more effectively than flat file formats such as `.csv`. This is particularly useful for analytically work.
```{r}
healthy_wave=haven::read_sav(here("data","CRONOS3 Theme - Make it Healthy.sav"))
dictionary=labelled::generate_dictionary(healthy_wave)
```

# Exploratory Data Analysis (EDA)

## Duplicate value 

```{r}
sum(duplicated(healthy_wave$idno))
```

```{r}

healthy_wave$needed_any=ifelse(
  rowSums(healthy_wave[, c("w1hq1", "w1hq3", "w1hq6", "w1hq11")] == 1, na.rm = TRUE) > 0,
  1, 0
)
```

```{r}
healthy_wave %>%
  filter(idno==25213)|>
  select(cntry,age,idno)|>
  haven::as_factor() |>
  tt()
```
Checking id-25213  

In `healthy_wave1`, 25213 id has two entries with countries as CZ and FI.

Trying to see how the ids are filled, realized that for every country there is a given range of values assigned.

So as to create unique values, we will be merging `cntry` and `idno` to create a new value, which will be the new primary key for filtering.

Solution-1
solving duplicates by merging the `cntry` and `idno` column 

```{r}
primary_health_wave=healthy_wave|>
  mutate(new_id=paste(cntry,idno,sep= "_"))
sum(duplicated(primary_health_wave$new_id))
```
still 21 duplicate values exist so we merge another column which is `age` to create a primary key in the dataset
also to check if its not just the same person and not at different ages / years we had to check this 


```{r}
primary_health_wave|>
  filter(duplicated(new_id) | duplicated(new_id,fromLast=TRUE)) |>
  arrange(new_id)|>
  select(cntry, new_id,age,gndr)|>
  haven::as_factor()|>
  slice_head(n=12)|>
  tt()
```
from this we can say that none of them are the same person as there huge again difference gap of more  than one year 

on merging with with three columns which is  `age` ,`idno` and `cntry` we were able to remove the duplicates so now we can use this column as a primary key for identifying an individual

```{r}
primary_health_wave=primary_health_wave|>
  mutate(new_id=paste(cntry,idno,age, sep="_"))
sum(duplicated(primary_health_wave$new_id))
```



## Model parametrs at intial phase

 no of people who Needed to see a general practitioner last 12 months but could not 
```{r}
sum(healthy_wave$w1hq1 ==1,na.rm = TRUE)
```
no fo people who Needed to see dentist last 12 months but could not
```{r}
sum(healthy_wave$w1hq6 ==1,na.rm = TRUE)
```
no of people Needed to see medical specialists but could not 
```{r}
sum(healthy_wave$w1hq3 == 1, na.rm = TRUE)
```

```{r}
length(healthy_wave$idno[
  which(healthy_wave$w1hq6==1|
        healthy_wave$w1hq3==1|
        healthy_wave$w1hq1==1|
        healthy_wave$w1hq11==1)
])
```

we have increased the number of observation to 1048

### Variable Description Table

| Variable | Role | Description |
|----------|------|-------------|
| w1hq1 | Response | Needed to see a general practitioner in the last 12 months |
| w1hq3 | Response | Needed to see medical specialists in the last 12 months |
| w1hq6 | Response | Needed to see dentist in the last 12 months |
| w1hq11 | Response | Needed diagnostic tests or scans prescribed by a doctor in the last 12 months |
| w1hq5 | Predictor | Wait time for medical specialist appointment |
| cntry | Predictor | Country |
| age | Predictor | Age |
| gndr | Predictor | Gender |
| w3hq77 | Predictor | Currently in paid work |
| w3hq57 | Predictor | Subjective mental health in general |
| w3hq58 | Predictor | Subjective physical health in general |
| w3hq87 | Predictor | Difficulty taking an hour off for personal/family matters during work |
| w5hq18 | Predictor | Healthcare costs covered by system |
| w5hq11 | Predictor | Receive help and support from close contacts when needed |
| w5hq8 | Predictor | Severe financial difficulties in family during first 18 years of life |
| w5hq17 | Predictor | Out-of-pocket cost for most recent private mental health consultation |
| w5hq12 | Predictor | Felt lonely during the past week |



## Country wise analysis
```{r}
country_summary=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    total_people=n(),
    people_needed=sum(needed_any, na.rm = TRUE),
    proportion=people_needed/total_people
  )|>
  haven::as_factor() |>
  tt()
country_summary
```

```{r}
primary_health_wave$needed_any=factor(
  primary_health_wave$needed_any,
  levels = c(0, 1),
  labels = c("Met Health Care", "Unmet Health Care")
)
```

```{r}
primary_health_wave$gndr=factor(
  primary_health_wave$gndr,
  levels = c(1, 2),
  labels = c("Male", "Female")
)
```

```{r}
ggplot(primary_health_wave)+
  geom_mosaic(aes(x=product(needed_any,cntry),
                  fill=needed_any))+
  scale_x_productlist("Country") +
  scale_y_productlist("Medical Care Status") +
  scale_fill_manual(
    values = c("orange", "blue"),
    name = "Medical Care Status"
  ) +
  theme_minimal()+
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```


The mosaic plot suggests that, across all countries, most respondents reported having their medical care needs met. However, the proportion of unmet health care need varies by country, with relatively higher shares observed in countries such as **GB(United Kingdom)**, **PL(Poland)**, **PT(Portugal)**, and **FI(Finland)**. This indicates potential cross-country differences in access to or delivery of medical care services.
```{r}
country_summary1=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    people_needed=sum(needed_any == "Unmet Health Care", na.rm = TRUE),
  )|>
  arrange(desc(people_needed))|>
  haven::as_factor() |>
  tt()
  

country_summary1
```

```{r}
service_country=primary_health_wave|>
  group_by(cntry)|>
  summarise(
    GP=sum(w1hq1==1,na.rm=TRUE),
    Specialist=sum(w1hq3==1,na.rm=TRUE),
    Dentist=sum(w1hq6==1,na.rm=TRUE),
    Diagnostic_Test=sum(w1hq11==1,na.rm=TRUE))|>
  pivot_longer(
    cols=-cntry,
    names_to="Service",
    values_to="Count")
ggplot(service_country,aes(x=reorder(cntry,-Count),y=Count,fill=Service))+
  geom_bar(stat="identity")+
  labs(title="Unmet Healthcare Services by Country",
    x="Country",
    y="Count")+
  theme_minimal()+
  theme(axis.text.x=element_text(angle=45,hjust=1))
```




## Variable Analysis thorugh likert charts 
| Variable | Description |
|----------|-------------|
| w3hq57 | Subjective mental health in general |
| w3hq58 | Subjective physical health in general |
| w5hq12 | Felt lonely, how much of the time in the past week |

```{r}
label_summary <- tibble(
  Variable = c(
    "Subjective mental health in general",
    "Subjective physical health in general",
    "Felt lonely, how much of the time past week"
  ),
  Labels = c(
    paste(names(attr(healthy_wave$w3hq57, "labels")), collapse = ", "),
    paste(names(attr(healthy_wave$w3hq58, "labels")), collapse = ", "),
    paste(names(attr(healthy_wave$w5hq12, "labels")), collapse = ", ")
  )
)

tt(label_summary)
```


The labels Very Positive, Positive, Decent, Negative, and Very Negative were chosen to create a consistent ordinal framework across the three well being variables: `loneliness`, `subjective mental health`, and `subjective physical health`. Although the original survey items use slightly different response wording (for example, frequency-based responses for loneliness and quality-based responses for health status), recoding them into a common directional scale improves the comparability and interpret ability of the Likert chart.

```{r}
primary_health_wave$lonely=factor(primary_health_wave$w5hq12,levels = c(1, 2, 3, 4),labels = c("Very Positive","Positive","Decent","Negative" ),ordered = TRUE)

primary_health_wave$mental_state=factor(primary_health_wave$w3hq57,levels = c(1, 2, 3, 4,5),labels = c("Very Positive","Positive","Decent","Negative","Very Negative"),ordered = TRUE)
primary_health_wave$physical_state=factor(primary_health_wave$w3hq58,levels = c(1, 2, 3, 4,5),labels = c("Very Positive","Positive","Decent","Negative","Very Negative"),ordered = TRUE)

df_yes=primary_health_wave[primary_health_wave$needed_any== "Unmet Health Care", ]
df_no=primary_health_wave[primary_health_wave$needed_any=="Met Health Care", ]
```


```{r}
plot_likert(df_yes[,c("mental_state", "physical_state","lonely")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
show.prc.sign = TRUE,geom.colors = "RdBu")
```

```{r}
plot_likert(df_no[,c("mental_state", "physical_state","lonely")],
grid.range = c(1.2, 1.4),
expand.grid = FALSE,
values = "sum.outside",
show.prc.sign = TRUE,geom.colors = "RdBu")
```
From this we can say that mental state of people who did receive medical care they had bad mental state as well as a bad physical state which makes it a good parameter for the model, we can also say loneliness is kind of high for people who need health care 


## Missing Data Analysis

| Variable Code | Description |
|--------------|-------------|
| w3hq87 | How difficult to take an hour off for personal/family matters during work |
| w5hq18 | Healthcare costs covered by system |
| w5hq11 | Receive help and support from close contacts when needed |
| w5hq8 | Severe financial difficulties in family during first 18 years of life, how often |
| w5hq17 | Out-of-pocket cost for most recent private mental health consultation |
| w5hq12 | Felt lonely, how much of the time past week |
| w3hq77 | Currently in paid work |


```{r}
missing_raw=miss_var_summary(
  df_yes[, c("w3hq87", "w5hq18", "w5hq11", "w5hq8", "w5hq17", "w5hq12","w3hq77")]
)
missing_table=tibble(
  Variable=c(
    "Time off work difficulty",
    "Healthcare cost coverage",
    "Social support",
    "Childhood financial difficulty",
    "Mental health consultation cost",
    "Loneliness",
    "Currently in paid work"
  ),
  Missing_Count=as.numeric(missing_raw$n_miss),
  Missing_Percentage=as.numeric(missing_raw$pct_miss)
)
tt(missing_table)
```
we might need to remove two varible for the model to take btter resutls 
w5hq17- Out-of-pocket cost for most recent private mental health consultation -86.6%
w3hq87- How difficult to take an hour off for personal/family matters during work -50.7% 
as they very high missing values so its better to remove them 


```{r}
gg_miss_upset(primary_health_wave[,c("w3hq87", "w5hq18","w5hq11","w5hq8","w5hq17","w5hq12","w3hq77")])
```



removing both those column remove a high chunk of missing values as we can see `w5hq17` and a combination of both `w3hq87` with `w5hq17` have the most missing values 



```{r}
gg_miss_case(primary_health_wave[,c("w3hq87", "w5hq18","w5hq11","w5hq8","w5hq17","w5hq12","w3hq77")])
```
comparing the miss cases on removing the two columns 
```{r}
gg_miss_case(primary_health_wave[,c( "w5hq18","w5hq11","w5hq8","w5hq12","w3hq77")])
```



```{r}
ggplot(primary_health_wave,aes(x=factor(needed_any),y=age,fill=factor(gndr)))+
  geom_boxplot(outlier.shape = NA)+ 
  labs(x="Healthcare Need Status",
    y="Age",
    title="Age Distribution by Healthcare Need Status")+
  scale_x_discrete(labels=c("Met","Unmet"))+
  scale_fill_manual(name="Gender",
    labels=c("Male", "Female","No answer"),
    values=c("blue","pink","black")
  )+facet_wrap(vars(cntry))
```

```{r}
primary_health_wave |>
  filter(!is.na(gndr))|>
  group_by(cntry, needed_any, gndr)|>
  summarise(
    n = n(),
    median_age = median(age, na.rm = TRUE),
    age_range = paste(
      quantile(age, 0.25, na.rm = TRUE),
      " - ",
      quantile(age, 0.75, na.rm = TRUE)
    ),.groups = "drop")|>
  haven::as_factor() |>
  tt()
```
Across most countries, the median ages for respondents with met and unmet healthcare needs are broadly similar, though some notable differences emerge.

- **Ireland:** unmet males = 59.5, met males = 57  
- **Austria:** unmet females = 56.5, met females = 53  
- **Poland:** unmet females = 49.5, met females = 45  

This suggests that in some countries, those reporting unmet healthcare needs tend to be slightly older. However, this is not universal.

- **Iceland:** unmet females = 42 vs met females = 54  
- **Slovenia:** unmet males = 40 vs met males = 52  

Across nearly all countries, female respondents generally have slightly lower median ages than males within the same healthcare need group. This pattern is visible in both the boxplot medians and the table.

The interquartile ranges (`age_range`) show that age dispersion differs across countries.

- **Austria** females unmet: 36–73.25 — wide spread  
- **Poland** females unmet: 36.5–68.25 — wide spread  
- **Iceland** females unmet: 31–52 — much narrower

$$
\begin{aligned}
\log \left(
\frac{P(\text{Unmet Health Care Need})}
{1 - P(\text{Unmet Health Care Need})}
\right)
&= \beta_0 + \beta_1(\text{Age}) + \beta_2(\text{Gender}) + \beta_3(\text{Country}) \\
&\quad + \beta_4(\text{Subjective Mental Health}) + \beta_5(\text{Subjective Physical Health}) \\
&\quad + \beta_6(\text{Loneliness}) + \beta_7(\text{Social Support}) \\
&\quad + \beta_8(\text{Currently in Paid Work}) \\
&\quad + \beta_9(\text{Early-Life Financial Difficulty}) \\
&\quad + \beta_{10}(\text{Healthcare Costs Covered})
\end{aligned}
$$



```{r}
model_age=glm(
  needed_any ~ age + gndr + cntry +
    w3hq57 + w3hq58 +w5hq12 + w5hq11 +w3hq77+w5hq8 + w5hq18 ,
  family = binomial,
  data = primary_health_wave
)

summary(model_age)
```



